# Formatting plots with `pubtheme`


The plots in these notes use the package `pubtheme`, which contains a `ggplot` theme `theme_pub` and a helper function `pub` for creating data journalism-style data visualizations with color palettes and formatting similar to those used by media organizations like BBC, NY Times, and ESPN. Several templates for scatter plot, line plots, etc., are provided for easy copying/pasting at the package's Github page, https://github.com/bmacGTPM/pubtheme. Installation instructions are at that page as well. 


## Simple Example

As an example, `pubtheme` turns a `ggplot` like this

```{r fig.height = 6, fig.width = 7, class.source = 'fold-hide'}
library(tidyverse)
library(pubtheme)

restore.ggplot.defaults()

ggplot(mtcars, 
       aes(x = wt, 
           y = mpg, 
           color = as.factor(cyl))) +
  geom_point(aes(size = mpg)) + 
  labs(title = 'Miles Per Gallon vs Weight', 
       x     = 'Weight', 
       y     = 'Miles Per Gallon',
       color = 'Cylinders', 
       size  = 'MPG') 

```

into this

```{r fig.height=7, fig.width=6}
dg = mtcars %>% 
  select(wt, mpg, cyl)

g = ggplot(mtcars, 
           aes(x = wt, 
               y = mpg, 
               color = as.factor(cyl)))+
  geom_point(aes(size = mpg))+
  labs(title = 'Miles Per Gallon vs Weight', 
       x     = 'Weight',  
       y     = 'Miles Per Gallon',
       color = 'Cylinders', 
       size  = 'MPG') 

g %>% 
  pub(xlim = c(0, 6),
      ylim = c(0, 40))
```

## Gallery 

Gallery is coming soon! For easy browsing, we will create a gallery of tables and visualizations along with hyperlinks to where each table or visualization is located in this book. 

For now, you can see several examples and templates on the Github page
https://github.com/bmacGTPM/pubtheme.

