# Schedule data

Let's investigate a few aspects of schedule data, like how often teams play each other, how many days rest teams typically get, and how game outcomes are or are not affected by how many days rest teams get before playing their games. 

## Matchups grid plot

Let's visualize how often teams play each other in a season. 

```{r fig.width=6, fig.height=7}
dg = d %>% 
  filter(season==2022) %>%
  group_by(away, home) %>%
  summarise(games = n()) %>%
  ungroup() %>%
  complete(away, home, fill=list(games=0))
head(dg)

title = "Number of Games Between Each Pair of Team" 
g = ggplot(dg, aes(x=home, y=away, fill=games))+
  geom_tile(linewidth=0.4, show.legend = F) +
  scale_fill_gradient(low = pubbackgray,
                      high = pubred,
                      na.value = 'white',
                      oob=squish) +
  labs(title    = title,
       x = 'Home Team', 
       y = 'Away Team')+  ## Optional. 
  geom_vline(xintercept=1:(length(unique(dg$home ))+1)-.5, color=pubdarkgray, linewidth=0.2)+ # vert  lines between each square
  geom_hline(yintercept=1:(length(unique(dg$away))+1)-.5, color=pubdarkgray, linewidth=0.2)+ # horiz lines 
  scale_x_discrete(expand = c(0, 0), position='top')+
  scale_y_discrete(expand = c(0, 0)) +
  theme_pub(type='grid', base_size=36/3)+
  theme(#axis.text.x = element_text(angle=90, vjust=-.5, hjust=0, margin())
        axis.text.x.top = element_text(
          angle = 90,
          margin= margin(b =  1/1440 * 20 * 12/36, unit = "in"),
          vjust = .5, 
          hjust = 0 ))

print(g)
```


## Order character vectors using `factor`

By default, R organizes the teams alphabetically. Instead of seeing teams alphabetically, we'll want to see them organized by division. So we'll use `factor` and specify the `levels` to be the order we want for both teams and divisions

```{r}
tms = read.csv('data/nba.teams.csv')
tms = tms %>% 
  arrange(conf, div) %>%
  mutate(conf = paste0(toupper(substr(conf,1,1)), substr(conf, 2, nchar(conf))),
         div  = paste0(toupper(substr(div ,1,1)), substr( div, 2, nchar( div))), 
         div  = factor(div, levels=unique(div)))
head(tms)

teams.order = tms %>% 
  select(team) %>% 
  unlist()
teams.order

dg = d %>% 
  filter(season==2022) %>%
  group_by(away, home) %>%
  summarise(games = n()) %>%
  ungroup() %>%
  complete(away, home, fill=list(games=0)) %>%
  left_join(tms %>% select(team, div), 
            by = c('home'='team')) %>%
  left_join(tms %>% select(team, div), 
            by = c('away'='team'), suffix=c('.h', '.a')) %>%
  mutate(away = factor(away, levels=rev(teams.order)), 
         home = factor(home, levels=teams.order))

head(dg)
```
## Matchups grid plot by division

```{r fig.width=6, fig.height=7}
title = "Number of Games Between Each Pair of Team" 
g = ggplot(dg, aes(x=home, y=away, fill=games))+
  geom_tile(linewidth=0.4, show.legend = F) +
  facet_grid(div.a ~ div.h, drop = T, scales='free')+
  scale_fill_gradient(low = pubbackgray,
                      high = pubred,
                      na.value = 'white',
                      oob=squish) +
  labs(title    = title,
       x = 'Home Team', 
       y = 'Away Team')+  ## Optional. 
  geom_vline(xintercept=1:(length(unique(dg$home ))+1)-.5, color=pubdarkgray, linewidth=0.2)+ # vert  lines between each square
  geom_hline(yintercept=1:(length(unique(dg$away))+1)-.5, color=pubdarkgray, linewidth=0.2)+ # horiz lines 
  scale_x_discrete(expand = c(0, 0), position='top')+
  scale_y_discrete(expand = c(0, 0)) +
  theme_pub(type='grid', base_size=36/4)+
  theme(axis.text.x.top = element_text(angle = 90, vjust = .5, hjust = 0), 
        panel.spacing = unit(10 * 1/1440 * 20 * 9/36, "in"))

print(g)
```

## Days rest by team

```{r}
dd = dd %>%
  arrange(date) %>%
  group_by(team) %>%
  mutate(days.rest = c(NA, diff(date))) 
dd %>% filter(team=='PHI') %>% head()
```
```{r}
dg = dd %>%
  filter(season==2022) %>%
  mutate(days.rest = ifelse(days.rest>=3, '3 or more', days.rest)) %>%
  group_by(team, days.rest) %>%
  summarise(games = n()) %>% 
  group_by(days.rest) %>%
  mutate(max = max(games))
  
head(dg)
```

```{r fig.height=6, fig.width=6}
title = "Number of Games With Each Number of Days Rest" 

g = ggplot(dg, aes(x=games, y=team))+
  geom_bar(stat='identity', aes(x=max+10), color=NA, fill=publightgray, width=0.8)+ ## optional background bars. 
  geom_bar(stat='identity', fill=pubred, color=NA, width=0.8)+ 
  geom_text(aes(label=round(games,2)), hjust=-0.1, vjust=0.4)+ ## optional numbers with reasonable number of digits
  facet_wrap(~days.rest, scales='free')+
  labs(title    = title,
       x = NULL, ## Optional. 
       y = NULL)+  ## Optional. Upper Lower.
  #scale_x_continuous(limits=c(0,dg$max))+ 
  theme_pub(type='bar', base_size=36/4) 
print(g)
```


## Days rest and game outcomes
Let's explore the relationship between `days.rest` and game outcomes. 

```{r}
dr = dd %>%
  group_by(days.rest) %>%
  summarise(score = round(mean(score),2), 
            opp.score = round(mean(opp.score),2), 
            n = n()) %>%
  mutate(days.rest = factor(days.rest, levels=6:1)) %>%
  filter(!is.na(days.rest))
dr
```



```{r}
dr = dr %>% 
  mutate(max=150, 
         label = paste0(round(score,1), ' (n = ', n, ')'))

title = "Average Points Scored by Days Rest" 
g = ggplot(dr, aes(x=score, y=days.rest))+
  geom_bar(stat='identity', aes(x=max), color=NA, fill=publightgray, width=0.8)+ ## optional background bars. 
  geom_bar(stat='identity', fill=pubred, color=NA, width=0.8)+ 
  geom_text(aes(label=label), hjust=-0.1)+ ## optional numbers with reasonable number of digits
  labs(title    = title,
       x = 'Average Points Scored', ## Optional. 
       y = 'Days Rest')+  ## Optional. Upper Lower.
  scale_x_continuous(limits=c(0,150))+ 
  theme_pub(type='bar', base_size=36/3) 
print(g)
```
```{r}
dr = dr %>% 
  mutate(max=150, 
         label = paste0(round(opp.score,1), ' (n = ', n, ')'))

title = "Average Points Allowed by Days Rest" 
g = ggplot(dr, aes(x=opp.score, y=days.rest))+
  geom_bar(stat='identity', aes(x=max), color=NA, fill=publightgray, width=0.8)+ ## optional background bars. 
  geom_bar(stat='identity', fill=pubred, color=NA, width=0.8)+ 
  geom_text(aes(label=label), hjust=-0.1)+ ## optional numbers with reasonable number of digits
  labs(title    = title,
       x = 'Average Points Allowed', ## Optional. 
       y = 'Days Rest')+  ## Optional. Upper Lower.
  scale_x_continuous(limits=c(0,150))+ 
  theme_pub(type='bar', base_size=36/3) 
print(g)
```
