# (PART) Data exploration and visualization {-} 

# Explore EV charging station data

## Lesson Objectives

Discussion

- What questions about EV charging stations are you interested in answering? 
- Introduce the data
- How can you answer those with this data? Can some of them not be answered? Possibly update your list of questions. 
- Live coding. We’ll start answering those questions in class. 

## Get experience with dplyr 

- filter, select, mutate, rename, summarize, arrange, group_by, *_join, pivot_longer, pivot_wider, ifelse, case_when
- String multiple commands together with pipe operator

## Get experience with ggplot and other plotting tools

- geom_’s: point, jitter, smooth, line, bar, hist, tile, segment
- Introduce the `pubtheme` package
- Faceting
- `ggplotly`
- `DT`  


## EV Charging Stations data


```{r nice-fig2, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab2, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].
