# Maps

## Stations in US and Canada

```{r}
title = "Level 2 and Level 3 Charging Stations in US and Canada"
g = ggplot(d, aes(x=lon, y=lat, size=lev23))+
  borders('state', colour=publightgray)+
  geom_point(alpha=.1, color=pubred, show.legend = F)+
  labs(title    = title)+
  coord_cartesian(expand=F)+
  theme_pub(type='map', base_size = 36/3)
g
```

## Level 2 and Level 3 stations

```{r fig.width=8, fig.height=3.5}
dg = d %>% 
  pivot_longer(cols = c(lev2, lev3)) %>%
  filter(value!=0)

title = "Level 2 and Level 3 Charging Stations in US and Canada"
g = ggplot(dg, aes(x=lon, y=lat, size=value))+
  borders('state', colour=publightgray)+
  geom_point(alpha=.1, color=pubred, show.legend = F)+
  facet_wrap(~name)+
  labs(title    = title)+
  coord_cartesian(expand=F)+
  theme_pub(type='map', base_size = 36/3)
g
```

## Level 2 Charging Stations by Network
```{r fig.height=5, fig.width=8}
dg = d %>%
  filter(lev2 > 0) %>% 
  filter(status=='avail')

title = "Level 2 Charging Station Locations in US and Canada"
g = ggplot(dg, aes(x=lon, y=lat, color=network2, size=lev2))+
  borders('usa', colour=publightgray)+
  geom_point(alpha=.2, show.legend = F)+
  facet_wrap(~network2)+
  labs(title    = title, 
       caption  = 'As of Jan 15, 2023. Data: https://afdc.energy.gov/')+
  #lims(x=c(-130,-62), y=c(23, 51))+
  coord_cartesian(expand=F)+
  scale_color_manual(
    values = c('orange', pubdarkgray, pubred,  ## let's use approximate
               'forestgreen', pubblue, pubmediumgray))+     ## company colors
  scale_size(range=c(.75,3))+
  theme_pub(type='map', base_size = 36/3)+
  theme(legend.justification = 'center')#+ 
  #guides(color=guide_legend(nrow=2,byrow=TRUE, hjust=0.5), size='none')
g
```


## Level 3 Charging Stations by Network

```{r fig.height=5, fig.width=8}
dg = d %>%
  filter(lev3 > 0) %>% 
  filter(status=='avail')

title = "Level 3 Charging Station Locations in US and Canada"
g = ggplot(dg, aes(x=lon, y=lat, color=network3, size=lev3))+
  borders('usa', colour=publightgray)+
  geom_point(alpha=.3, show.legend = F)+
  facet_wrap(~network3)+
  labs(title    = title, 
       caption  = 'As of Jan 15, 2023. Data: https://afdc.energy.gov/')+
  #lims(x=c(-130,-62), y=c(23, 51))+
  coord_cartesian(expand=F)+
  scale_color_manual(
    values = c(pubred  , 'darkturquoise', 'navy',          ## let's use approx
               'darkorange', pubdarkgray, pubmediumgray))+ ## company colors
  scale_size(range=c(.75,3))+
  theme_pub(type='map', base_size = 36/3)+
  theme(legend.justification = 'center')#+ 
g
```


## Level 2 Charging Stations Animation (coming soon)

## Level 3 Charging Stations Animation (coming soon)

## Interative map with Leaflet

```{r}
dg = d %>% 
  filter(state=='CT') %>%
  mutate(label=paste0(network, '<br>', 
                      station.name, '<br>',
                      street.address, '<br>', 
                      city, ', ', state, ' ', zip, '<p>', 
                      'Level 2: ', lev2, '<br>',
                      'Level 3: ', lev3))
head(dg,2)
```

```{r fig.height=7, fig.width=9}
leaflet(data=dg, height=540, width=720) %>% 
  addTiles() %>% ## adds the background map
  addCircleMarkers(lng = ~lon,      ## add points 
                   lat = ~lat,
                   radius = 5,
                   color  = 'black',
                   stroke = FALSE, ## remove the border
                   fillOpacity = 0.5,
                   label = ~lapply(label, HTML), ## show a label when you hover
                   popup = ~label) %>% ## show a label when you click
  setView(lng=-72.79458, lat=41.51979, zoom=9) 
```

This is an interactive map with `leaflet`. There are labels that show the EV network, station name, street address, city, state, zip, and the number Level 2 and Level 3 charging stations. The dots are sized by the number of chargers at that location. Since our computer gets slow showing all the points, we are plotting just the locations in CT. 

## Plotting stations over census data

```{r fig.height=7, fig.width=10}
dc = readRDS('data/tracts.census.2019.CT.rds')
dc = dc[dc$state=='CT' & dc$pop!=0,]


## define a color palette
pal1 = colorQuantile(palette = c(pubbackgray, 
                                 pubblue), 
                     domain=NULL,
                     n=20);

## Create a label
dc@data$label = paste0('GEOID: ', 
                      dc$GEOID, '<br/>', 
                      dc$county, ', ',
                      dc$state, '<br/>', 
                      'Median Age: ', dc$house.value, '<br/>', 
                      'Population: ', comma(d$pop))

## leaflet
l1 = leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addTiles() %>%
  addPolygons(data = dc[dc$pop!=0,],
              fillColor = ~pal1(pop.density),
              label= ~label %>% lapply(HTML),
              fillOpacity = 0.9,
              color = 'black',
              weight = 1) %>%
  addCircleMarkers(data=dg[dg$lev2!=0,],
                   lng = ~lon,       
                   lat = ~lat,
                   radius=~lev2, 
                   color=pubdarkgray, 
                   label=~label %>% lapply(HTML)) %>%
  addCircleMarkers(data=dg[dg$lev3!=0,],
                   lng = ~lon,      
                   lat = ~lat,
                   radius=~lev3, 
                   color=pubred, 
                   label=~label %>% lapply(HTML)) %>%
  setView(lng=-72.79458, lat=41.51979, zoom=9) 
  
l1
```

