# Data visualization with `ggplot`

First let's prep our data set using our code from the last section. 
```{r}
d = readRDS('data/games.rds')
d = d %>% 
  filter(lg=='nba', 
         season %in% 2022, 
         season.type=='reg') %>%
  select(date, away, home, ascore, hscore, season, gid) 

da = d %>% select(date, away, ascore, home, hscore, season, gid) %>% mutate(ha = 'away')
dh = d %>% select(date, home, hscore, away, ascore, season, gid) %>% mutate(ha = 'home')
cols = c('date', 'team', 'score',  'opp', 'opp.score', 'season', 'gid', 'ha')
colnames(da) = cols
colnames(dh) = cols

dd = bind_rows(da, dh) %>% 
  arrange(date) 
head(dd,2)
```


## Scatter plot with `geom_point`

```{r fig.width=5, fig.height=5}
ggplot(d, aes(x=ascore, y=hscore))+
  geom_point()
```

## Trend line with `geom_smooth`

## Scatter plot with `geom_jitter`

When there are a lot of overlapping points in a plot, it can help to use `geom_jitter` in place of `geom_point`.

```{r fig.height=6, fig.width=6}
ggplot(d, aes(x=hscore, y=home))+
  geom_point()
```

That has a lot of points plotted on top of one another. `geom_jitter` adds randomness to the points so that don't overlap as much. Here we add a little randomness in the vertical direction, and none on the horiztonal direction. 
```{r fig.height=6, fig.width=6}
ggplot(d, aes(x=hscore, y=home))+
  geom_jitter(height=0.2, width=0)
```


## Bar plot `geom_bar`


```{r}
dg = dd %>% 
  group_by(team) %>%
  summarise(score=mean(score), 
            opp.score = mean(opp.score))
head(dg)
```
```{r}
ggplot(dg, aes(x=score, y=team))+
  geom_bar(stat='identity')
```

## Text with `geom_text`



## Histogram with `geom_histogram`
Histogram of points scored
```{r}
ggplot(dd, aes(x=score))+
  geom_histogram(color='black', binwidth = 5)
```

## Line plot with `geom_line`
```{r}
ggplot(dd %>% filter(team=='PHI'), aes(x=date, y=score))+
  geom_line()
```


## Grid plot with `geom_tile`

Let's show how often teams play against one another. 
```{r}
dg = d %>%
  group_by(home, away) %>%
  summarise(games = n())
head(dg)
```

```{r fig.width=6, fig.width=6}
ggplot(dg, aes(x=home, 
               y=away, 
               fill=as.numeric(games), 
               color='black')) + 
  geom_tile()

df <- data.frame(
  x = rep(c(2, 5, 7, 9, 12), 2),
  y = rep(c(1, 2), each = 5),
  z = factor(rep(1:5, each = 2)),
  w = rep(diff(c(0, 4, 6, 8, 10, 14)), 2)
)
ggplot(df, aes(x, y)) +
  geom_tile(aes(color = z, fill = w))

```

## Faceting with `facet_wrap`

We can make separate plots based on one of the columns in the data. Here we make separate histograms for home and away teams (`ha`) and stack them on top of each other by selecting `rows=2`. 

```{r}
ggplot(dd, aes(x=score))+
  geom_histogram(binwidth=5, color='black')+
  facet_wrap(~ha, nrow=2)
```


## For more information

The book [R for Data Science](https://r4ds.had.co.nz/) is a great resource for `tidyverse`. 

The `tidyverse` [website](https://www.tidyverse.org/) has links to main pages for `tidyverse` packages (e.g. `dplyr` [vignette](https://dplyr.tidyverse.org/) and `ggplot2` [vignette](https://ggplot2.tidyverse.org/)). 

Posit (formerly R Studio) has a [collection of cheetsheets](https://posit.co/resources/cheatsheets/) for these packages. 