# Data visualization with `ggplot`

First let's prep our data set using our code from the last section. 
```{r}
d = readRDS('data/games.rds')
d = d %>% 
  filter(lg=='nba', 
         season %in% 2022) %>%
  select(date, away, home, ascore, hscore, season, season.type, gid) %>%
  mutate(season.type = factor(season.type, levels=c('reg', 'post')))

da = d %>% select(date, away, ascore, home, hscore, season, season.type, gid) %>% mutate(ha = 'away')
dh = d %>% select(date, home, hscore, away, ascore, season, season.type, gid) %>% mutate(ha = 'home')
cols = c('date', 'team', 'score',  'opp', 'opp.score', 'season', 'season.type', 'gid', 'ha')
colnames(da) = cols
colnames(dh) = cols

dd = bind_rows(da, dh) %>% 
  arrange(date) 
head(dd,2)
```


## Scatter plot with `geom_point`

```{r fig.width=5, fig.height=5}
ggplot(d, aes(x=ascore, y=hscore))+
  geom_point()
```

## Color, size, shape

We can color by any variable in the data set by putting `color=variable` inside `aes()`. In this case, we’ll choose the categorical variable `season.type`.
```{r}
ggplot(d, aes(x=ascore, y=hscore, color=season.type))+
  geom_point()
```
If we want to change the color of all points to the same color, we can specify that color inside `geom_point()` instead.

```{r}
ggplot(d, aes(x=ascore, y=hscore))+
  geom_point(color='gray')
```
Likewise, we can size each point, or change the shape of each point, based on column in the data. 

```{r}
ggplot(d, aes(x=ascore, y=hscore, 
              color = season.type, 
              size = season.type, 
              shape = season.type))+
  geom_point()
```

(You normally wouldn't choose all three to be the same column, this is just for illustration.)

## Trend line with `geom_smooth`

We can add a trendline to our scatter plot using `geom_smooth` after `geom_point`. 
```{r}
ggplot(d, aes(x=ascore, y=hscore))+
  geom_point()+
  geom_smooth()
```

By default it will be a smooth curve as opposed to a line. For a line, use `method='lm'` (`lm` is the function used for linear regression).

```{r}
ggplot(d, aes(x=ascore, y=hscore))+
  geom_point()+
  geom_smooth(method='lm')
```


## Grouping observations with `group`

If you use color, then you get separate trendlines for each group.
```{r}
ggplot(d, aes(x=ascore, y=hscore, color=season.type))+
  geom_point()+
  geom_smooth(method='lm')
```
If group isn’t specified, ggplot assumes it should group by color. If we want to group by something other than what we are coloring by, we can specify it. Or we can say NA to say don’t group and get a single line.

```{r}
ggplot(d, aes(x=ascore, y=hscore, color=season.type, group=NA))+
  geom_point()+
  geom_smooth(method='lm')
```

## Faceting with `facet_wrap`

For categorical variables, we can create a separate window for each value of that variable. Here we make separate scatter plots for each season.type, `reg` and `post`. Note that if we use `geom_smooth`, there’s a line for each window and color.

```{r}
ggplot(d, aes(x=ascore, y=hscore, color=season.type))+
  geom_point()+
  geom_smooth(method='lm')+
  facet_wrap(~season.type)
```

## Transparency using `alpha`
If there are lots of overlapping points, and we can adjust the transparency using `alpha`.

First here is a plot using `geom_point` that has a lot of points plotted on top of one another.

```{r fig.height=6, fig.width=6}
ggplot(d, aes(x=hscore, y=home))+
  geom_point()
```

Now we add `alpha=0.5`, which means we want points to be halfway between transparent and opaque (1=opaque, 0=transparent).

```{r fig.height=6, fig.width=6}
ggplot(d, aes(x=hscore, y=home))+
  geom_point(alpha=0.5)
```

## Scatter plot with `geom_jitter`

When there are a lot of overlapping points in a plot, it can also help to use `geom_jitter` in place of `geom_point`. 

We use `geom_jitter` to add randomness to the points so that don't overlap as much. Here we add a little randomness in the vertical direction, and none on the horizontal direction. We include `alpha=0.5` for some extra help.

```{r fig.height=6, fig.width=6}
ggplot(d, aes(x=hscore, y=home))+
  geom_jitter(alpha=0.5, height=0.2, width=0)
```

## Bar plot `geom_bar`

We can make bar plots using `geom_bar` like this. 

```{r}
ggplot(dd, aes(x=score, y=team))+
  geom_bar(stat='identity')
```
That counts the number of point score by each team. Note that we put `stat='identity'` inside `geom_bar()`. 

## Color bars using `fill`
We can also use color our bars using the `fill` aesthetic. 

```{r}
ggplot(dd, aes(x=score, y=team, fill=season.type))+
  geom_bar(stat='identity')
```

Typically you use `fill`. Note that `color` affects only the border.

```{r}
ggplot(dd, aes(x=score, y=team, color=season.type))+
  geom_bar(stat='identity')
```

That's not ideal. You may want to aggregate the data first like this

```{r}
dg = dd %>%
  group_by(team, season.type) %>%
  summarise(score = sum(score))
ggplot(dg, aes(x=score, y=team, color=season.type))+
  geom_bar(stat='identity')
```
But again, you'll typically use `fill`. 

## Side-by-side bar charts with `position_dodge`
Stacked bar are often not desired, so you can 
```{r}
## remove teams with no `post` data.
post.teams = unique(dg$team[dg$season.type=='post'])
dg = dg %>% filter(team %in% post.teams)
ggplot(dg, aes(x=score, y=team, fill=season.type))+
  geom_bar(stat='identity', position=position_dodge(), color='black')
```

## Text with `geom_text`


## Histogram with `geom_histogram`
Histogram of points scored
```{r}
ggplot(dd, aes(x=score))+
  geom_histogram(color='black', binwidth = 5)
```

We can use the `fill` aesthetic here too. 
```{r}
ggplot(dd, aes(x=score, fill=season.type))+
  geom_histogram(color='black', binwidth = 5)
```

## Line plot with `geom_line`
```{r}
ggplot(dd %>% filter(team=='PHI'), aes(x=date, y=score))+
  geom_line()
```


## Grid plot with `geom_tile`

Let's show how often teams play against one another. 
```{r}
dg = d %>%
  group_by(home, away) %>%
  summarise(games = n())
head(dg)
```

```{r fig.width=6, fig.width=6}
ggplot(dg, aes(x=home, 
               y=away, 
               fill=as.numeric(games), 
               color='black')) + 
  geom_tile()



```

## Quick interactive plots with `ggplotly`

We can make any of these plots interactive using `ggplotly`. By default, a tooltip is created that shows whatever aesthetics are used inside `aes()`. 
```{r}
g = ggplot(d, aes(x=ascore, y=hscore))+
  geom_point()
ggplotly(g)
```


## For more information

The book [R for Data Science](https://r4ds.had.co.nz/) is a great resource for `tidyverse`. 

The `tidyverse` [website](https://www.tidyverse.org/) has links to main pages for `tidyverse` packages (e.g. `dplyr` [vignette](https://dplyr.tidyverse.org/) and `ggplot2` [vignette](https://ggplot2.tidyverse.org/)). 

Posit (formerly R Studio) has a [collection of cheetsheets](https://posit.co/resources/cheatsheets/) for these packages. 