# Code tests and examples

## Scrollable code blocks test

Height of code. This changes for all code blocks?

```{css, echo=FALSE}
pre {
  max-height: 300px;
  overflow-y: auto;
}

pre[class] {
  max-height: 100px;
}
```

Height of output

```{css, echo=FALSE}
.scroll-100 {
  max-height: 100px;
  overflow-y: auto;
  background-color: inherit;
}
```



```{r}
# pretend that we have a lot of code in this chunk
if (1 + 1 == 2) {
  # of course that is true
  print(mtcars)
  # we just printed a lengthy data set
}
```

I assume this is how to change it for only one code chunk. 
```{r, class.output="scroll-100"}
print(mtcars)
## some comments
## so that 
## there is more code here
## to test scrolling
```

Note that those are both scrollable and hideable. 

## `DT` test

See https://rstudio.github.io/DT/ and https://rstudio.github.io/DT/options.html for options. 

```{r}
library(DT)
datatable(mtcars, 
          filter='top')
```

```{r}
#javascript code that can be included in a datatable
##to renumber the rownames to be 1 to n when sorting and filter
js <- c(
  "table.on('draw.dt', function(){",
  "  var PageInfo = table.page.info();",
  "  table.column(0, {page: 'current'}).nodes().each(function(cell,i){",
  "    cell.innerHTML = i + 1 + PageInfo.start;",
  "  });",
  "})")
```

DT with a bunch of options

```{r}
d = mtcars %>%
  mutate(perc = hp/max(hp)) ## add a dumb perc column
ddt = d %>%
  datatable(colnames = gsub('[.]', ' ', colnames(d)),
            rownames = T,
            filter = 'top',
            style = 'bootstrap',
            class = 'display nowrap compact', #'table-condensed table-hover table-responsive ', # table-nowrap didn't work
            extensions = c('ColReorder',  'FixedHeader', 'KeyTable', 'Buttons'),
            # escape=9, ## was for if you have HTML.  like HTML for the arrows
            options = list(
              colReorder = TRUE, ## formatRound stopped working. so gotta round manually above
              #scrollX = TRUE, fixedColumns = TRUE, ## alignment gets messed up
              fixedHeader=T,
              # keys = TRUE, don't like that
              dom = 'Brtip', ## each letter means something. Forget what.
              buttons = c('copy', 'csv', 'excel'),
              pageLength=1000,
              columnDefs=list(list(orderSequence = c('desc', 'asc'), targets = '_all' ))
               
            )#,
            #callback=JS(js) ## use the JS defined above
  )
ddt
# ddt = ddt %>% 
#   formatStyle('mpg', background = styleColorBar(mtcars$mpg, 'lightblue')) %>%
#   formatStyle('wt',
#               backgroundSize = '97% 90%',
#               backgroundRepeat = 'no-repeat',
#               backgroundPosition = 'center') %>%
#   formatPercentage('perc', 1)
# ddt
```

Unfortunately it doesn't show order-column. Compact doesn't work in bookdown, but does work in shiny for some reason.

## Shiny app

External

```{r}
knitr::include_app('https://bmac.shinyapps.io/CLTapp/')
```

In-line. Sadly this looks like it only takes a screen shot. It's not interactive. This can be done in R Markdown but the `bookdown` reference manual only talks about embedding external apps. 

Referencing datatables with links: https://stackoverflow.com/questions/49819892/cross-referencing-dtdatatable-in-bookdown

Styling options: https://datatables.net/manual/styling/classes

Reference: https://rstudio.github.io/DT/

```{r, echo=FALSE}
library(shiny)
shinyApp(

  ui = fluidPage(
    selectInput("region", "Region:",
                choices = colnames(WorldPhones)),
    plotOutput("phonePlot")
  ),

  server = function(input, output) {
    output$phonePlot = renderPlot({
      barplot(WorldPhones[,input$region]*1000,
              ylab = "Number of Telephones", xlab = "Year")
    })
  },

  options = list(height = 500)
)
```



## Stuff I want to remember

Code for selecting x range for checking normality and residuals in a linear regression model.
```
dragmode = 'select', 
selectdirection = 'h', ## for selecting an x range
```

## Stuff removed from _output.yml
```{r}
# bookdown::pdf_book:
#   includes:
#     in_header: preamble.tex
#   latex_engine: xelatex
#   citation_package: natbib
#   keep_tex: yes
# bookdown::epub_book: default
```
