# Random

This is a collection of random (in the non-statistical meaning of the word) stuff that I have found to be useful over the years. 

## Minimal reproducible examples

When asking someone for help in an email, on Slack, on Github, on a discussion board (e.g. stackoverflow.com), etc., use a minimal reproducible example. Here is the Stack Overflow page on the topic: https://stackoverflow.com/help/minimal-reproducible-example

- Minimal: Use as little code as necessary that still results in the error
- Reproducible: provide all code and data necessary so that someone can copy/paste and reproduce the problem on their own machine.  

Here is an example. Suppose you are working on a schedule matrix and have this code:

```{r eval=F}
d = readRDS('data/games.rds')
dg = d %>% 
  filter(lg=='nba', season %in% 2022, season.type=='reg') %>%
  group_by(away, home, season) %>%
  summarise(games = n()) %>%
  ungroup() %>%
  complete(away, home, fill=list(games=0))

head(dg)

title = "Number of Games Between Each Pair of Teams" 
g = ggplot(dg, aes(x=home, y=away, fill=games))+ 
  geom_tile(linewidth=0.4, show.legend = T, color=pubdarkgray) + ## used char above so leg is discrete
  scale_fill_manual(values=c(pubbackgray, publightred, pubred)) +
  labs(title    = title,
       x = 'Home Team', 
       y = 'Away Team')+  
  scale_x_discrete(expand = c(0, 0), position='top')+
  scale_y_discrete(expand = c(0, 0)) +
  theme_pub(type='grid', base_size=36/3)+
  theme(axis.text.x.top = element_text(angle = 90, vjust = .5, hjust = 0))
g
```

This results in the error `Error: Continuous value supplied to discrete scale`. Suppose we want help on this error. If the person we are asking has the `games.rds` data, then we can start stripping down the `ggplot` code as much as possible until we still have the error. Most lines of code can be removed because they are unrelated to the error. You can start commenting out one line at a time. This still gives the same error:

```{r eval=F}
title = "Number of Games Between Each Pair of Teams" 
g = ggplot(dg, aes(x=home, y=away, fill=games))+ 
  geom_tile(linewidth=0.4, show.legend = T, color=pubdarkgray) + 
  scale_fill_manual(values=c(pubbackgray, publightred, pubred)) #+
  # labs(title    = title,
  #      x = 'Home Team', 
  #      y = 'Away Team')#+  
  #scale_x_discrete(expand = c(0, 0), position='top')+
  #scale_y_discrete(expand = c(0, 0))# +
  #theme_pub(type='grid', base_size=36/3)#+
  #theme(axis.text.x.top = element_text(angle = 90, vjust = .5, hjust = 0))
g
```

So we can delete those lines of code. If we remove of `scale_fill_manual` the error goes away, so we have to keep that in. 

```{r eval=F}
g = ggplot(dg, aes(x=home, y=away, fill=games))+ 
  geom_tile(linewidth=0.4, show.legend = T, color=pubdarkgray) + 
  scale_fill_manual(values=c(pubbackgray, publightred, pubred)) 
g
```
 We can also try getting rid of some of the arguments, like `linewidth`, `show.legend`, `color`, and we still get the error. Also, we can change the colors `pubbackgray`, `publightred`, and `pubred` to `'gray`', `'lightpink'`, and `'red'` which come with `R` and don't require the `pubtheme` package. 

```{r eval=F}
g = ggplot(dg, aes(x=home, y=away, fill=games))+ 
  geom_tile() + 
  scale_fill_manual(values=c('gray', 'lightpink', 'red')) 
g
```

That is about all we can remove and still get the error. So if the person we are asking the question to has this data, they can copy/paste this code to their own computer and reproduce the error. 

If the person we are asking doesn't have the data, then we should use a widely available data set. The `mtcars` data is available to anyone with `R` and can be used here. 

```{r}
head(mtcars)
```

```{r eval=F}
library(tidyverse) 
dg = mtcars %>%
  group_by(cyl, vs) %>%
  summarise(mpg = mean(mpg), .groups='keep')
head(dg)

g = ggplot(dg, aes(x=cyl, y=vs, fill=mpg))+ 
  geom_tile() + 
  scale_fill_manual(values=c('gray', 'lightpink', 'red'))
g
```

Anyone, regardless of whether they have your data set, can copy/paste this code to diagnose. They will hopefully notice that `mpg`, the variable chosen to `fill` by, is continuous, while `scale_fill_manual` applies to discrete color scales. You'd have to use a continuous color scale with the continuous variable `mpg`. Something like this works: 

```{r}
library(tidyverse) 
dg = mtcars %>%
  group_by(cyl, vs) %>%
  summarise(mpg = mean(mpg), .groups='keep')
head(dg)

g = ggplot(dg, aes(x=cyl, y=vs, fill=mpg))+ 
  geom_tile() + 
  scale_fill_gradient(low='gray', high='red')
g
```

Now that we know how we can fix the error, let's go back to our original plot and replace `scale_fill_manual` with `scale_fill_gradient`. 

```{r fig.height=7, fig.width=7}
d = readRDS('data/games.rds')
dg = d %>% 
  filter(lg=='nba', season %in% 2022, season.type=='reg') %>%
  group_by(away, home, season) %>%
  summarise(games = n()) %>%
  ungroup() %>%
  complete(away, home, fill=list(games=0))
head(dg)

title = "Number of Games Between Each Pair of Teams" 
g = ggplot(dg, aes(x=home, y=away, fill=games))+ 
  geom_tile(linewidth=0.4, show.legend = T, color=pubdarkgray) + 
  scale_fill_gradient(low=pubbackgray, high=pubred) +
  labs(title    = title,
       x = 'Home Team', 
       y = 'Away Team')+  
  scale_x_discrete(expand = c(0, 0), position='top')+
  scale_y_discrete(expand = c(0, 0)) +
  theme_pub(type='grid', base_size=36/3)+
  theme(axis.text.x.top = element_text(angle = 90, vjust = .5, hjust = 0))
g
```

Another option would have been to convert `games` to a character or factor and keep `scale_fill_manual`. 



```{r fig.height=7, fig.width=7}
dg = dg %>%
  mutate(games = as.character(games))

title = "Number of Games Between Each Pair of Teams" 
g = ggplot(dg, aes(x=home, y=away, fill=games))+ 
  geom_tile(linewidth=0.4, show.legend = T, color=pubdarkgray) + ## used char above so leg is discrete
  scale_fill_manual(values = c(pubbackgray, publightred, pubred)) +
  labs(title    = title,
       x = 'Home Team', 
       y = 'Away Team')+  
  scale_x_discrete(expand = c(0, 0), position='top')+
  scale_y_discrete(expand = c(0, 0)) +
  theme_pub(type='grid', base_size=36/3)+
  theme(axis.text.x.top = element_text(angle = 90, vjust = .5, hjust = 0))
g
```


## Accessible Documents

Ways to improve accessibility of documents. 

- Use color-blind friendly color palettes.  `pubtheme` uses red/blue by default and contains `cb14`, a 14-color color blind friendly color palette.
- Add alt text to figures created in R Markdown https://www.rstudio.com/blog/knitr-fig-alt/
- You can knit to Word, click Review, and click Check Accessibility
- See https://bookdown.org/yihui/rmarkdown-cookbook/html-accessibility.html. That page also has a link to here  https://r-resources.massey.ac.nz/rmarkdown/.




## Math symbols in plots

This describes how to use mathematical expressions in plots. 

```
?plotmath
```

Examples are below. Some symbols may not appear correctly, but hopefully you'll get the idea. Note the use of `expression()` in the examples. For the symbols to appear, `expression()` must be used like this

```{r}
dg = mtcars %>% 
  select(wt, mpg, cyl) %>%
  mutate(Cylinders = as.factor(cyl)) %>%
  rename(MPG = mpg)

title = expression(Title~In~Upper~Lower~With~Symbols %=~% tilde(alpha))
g = ggplot(dg, aes(x=wt, y=MPG, color=Cylinders))+
  geom_point(aes(size=MPG))+
  labs(title    = title,
       subtitle = 'Optional Subtitle In Upper Lower',
       caption  = "Optional caption giving more info, Twitter handle, or shameless promotion of pubtheme",
       x = expression(x %+-% y),
       y = 'Vertical Axis Label in Upper Lower')+
  scale_x_continuous(limits=c(0, 6), breaks=c(0, 3, 6), oob=squish, labels=comma)+
  scale_y_continuous(limits=c(0,40), breaks=c(0,20,40), oob=squish, labels=comma)+
  coord_cartesian(clip='off', expand=FALSE)+
  scale_size(range=c(2,6))+
  theme_pub(type='scatter', base_size = 36/3) 
print(g)
```


```{r fig.width=8}
demo(plotmath)
```

