# Random

This is a collection of random (in the non-statistical meaning of the word) stuff that I have to be useful over the years. 

## Math symbols in plots

This describes how to use mathematical expressions in plots. 

```
?plotmath
```

Examples are below. Some symbols may not appear correctly, but hopefully you'll get the idea. Note the use of `expression()` in the examples. For the symbols to appear, `expression()` must be used like this

```{r}
dg = mtcars %>% 
  select(wt, mpg, cyl) %>%
  mutate(Cylinders = as.factor(cyl)) %>%
  rename(MPG = mpg)

title = expression(Title~In~Upper~Lower~With~Symbols %=~% tilde(alpha))
g = ggplot(dg, aes(x=wt, y=MPG, color=Cylinders))+
  geom_point(aes(size=MPG))+
  labs(title    = title,
       subtitle = 'Optional Subtitle In Upper Lower',
       caption  = "Optional caption giving more info, Twitter handle, or shameless promotion of pubtheme",
       x = expression(x %+-% y),
       y = 'Vertical Axis Label in Upper Lower')+
  scale_x_continuous(limits=c(0, 6), breaks=c(0, 3, 6), oob=squish, labels=comma)+
  scale_y_continuous(limits=c(0,40), breaks=c(0,20,40), oob=squish, labels=comma)+
  coord_cartesian(clip='off', expand=FALSE)+
  scale_size(range=c(2,6))+
  theme_pub(type='scatter', base_size = 36/3) 
print(g)
```


```{r fig.width=8}
demo(plotmath)
```

