# Code tests and examples

## Code folding test

Code folding test. This isn't working :(. Doesn't matter if `code_folding` is `show` or `hide` in the `_output.yml` file. Neither of these work. 

```{r class.source = "fold-show"}
head(mtcars)
```

```{r class.source = "fold-hide"}
head(mtcars)
```

## Scrollable code blocks test

Height of code. This changes for all code blocks?
  ```{css, echo=FALSE}
pre {
  max-height: 300px;
  overflow-y: auto;
}

pre[class] {
  max-height: 100px;
}
```

Height of output
```{css, echo=FALSE}
.scroll-100 {
  max-height: 100px;
  overflow-y: auto;
  background-color: inherit;
}
```



```{r}
# pretend that we have a lot of code in this chunk
if (1 + 1 == 2) {
  # of course that is true
  print(mtcars)
  # we just printed a lengthy data set
}
```

I assume this is how to change it for only one code chunk. 
```{r, class.output="scroll-100"}
print(mtcars)
## some comments
## so that 
## there is more code here
## to test scrolling
```

Note that those are both scrollable and hideable. 

## `DT` test



See https://rstudio.github.io/DT/ and https://rstudio.github.io/DT/options.html for options. 
```{r}
library(DT)
datatable(mtcars, 
          filter='top')
```

```{r}
#javascript code that can be included in a datatable
##to renumber the rownames to be 1 to n when sorting and filter
js <- c(
  "table.on('draw.dt', function(){",
  "  var PageInfo = table.page.info();",
  "  table.column(0, {page: 'current'}).nodes().each(function(cell,i){",
  "    cell.innerHTML = i + 1 + PageInfo.start;",
  "  });",
  "})")
```

DT with a bunch of options

```{r}
d = mtcars %>%
  mutate(perc = hp/max(hp)) ## add a dumb perc column
ddt = d %>%
  datatable(colnames = gsub('[.]', ' ', colnames(d)),
            rownames = T,
            filter = 'top',
            style = 'bootstrap',
            class = 'table-condensed table-hover table-responsive ', # table-nowrap didn't work
            extensions = c('ColReorder',  'FixedHeader', 'KeyTable', 'Buttons'),
            # escape=9, ## was for if you have HTML.  like HTML for the arrows
            options = list(
              colReorder = TRUE, ## formatRound stopped working. so gotta round manually above
              #scrollX = TRUE, fixedColumns = TRUE, ## alignment gets messed up
              fixedHeader=T,
              # keys = TRUE, don't like that
              dom = 'Brtip', ## each letter means something. Forget what.
              buttons = c('copy', 'csv', 'excel'),
              pageLength=1000,
              columnDefs=list(list(orderSequence = c('desc', 'asc'), targets = '_all' ))
            ),
            callback=JS(js) ## use the JS defined above
  )

ddt = ddt %>% 
  formatStyle('mpg', background = styleColorBar(mtcars$mpg, 'lightblue')) %>%
  formatStyle('wt',
              backgroundSize = '97% 90%',
              backgroundRepeat = 'no-repeat',
              backgroundPosition = 'center') %>%
  formatPercentage('perc', 1)
ddt
```



## Shiny app

```{r}
include_app('https://bmac.shinyapps.io/CLTapp/')
```

## Stuff removed from _output.yml
```{r}
# bookdown::pdf_book:
#   includes:
#     in_header: preamble.tex
#   latex_engine: xelatex
#   citation_package: natbib
#   keep_tex: yes
# bookdown::epub_book: default
```
