# Travel data

We'd like to explore the relationship between travel distance and game outcomes. First we need the distance between every pair of cities. Instead of the distance matrix format, we'll want one row per pair of cities, with columns location, previous locations, and miles between them. 

```{r}
library(tidyverse)
library(pubtheme)
d = readRDS('data/games.rds')
d = d %>% 
  filter(lg == 'nba', 
         season %in% 2021:2022, 
         season.type=='reg') %>%
  select(date, 
         away, home, 
         ascore, hscore, 
         season, gid)
head(d)

dm = read.csv('data/distance.matrix.nba.csv')
colnames(dm)[1] = 'prev.loc'
head(dm)

dm = dm %>% 
  pivot_longer(cols      = -prev.loc, 
               names_to  = 'loc', 
               values_to = 'miles')
head(dm)

tms = read.csv('data/nba.teams.csv')
tms = tms %>% 
  arrange(conf, div) %>%
  mutate(conf = paste0(toupper(substr(conf, 1, 1)), 
                               substr(conf, 2, nchar(conf))),
         div  = paste0(toupper(substr(div , 1, 1)), 
                               substr(div , 2, nchar( div))), 
         div  = factor(div, levels = unique(div)))
head(tms)


teams.order = tms %>% 
  select(team) %>% 
  unlist()
head(teams.order)

da = d %>% select(date, away, ascore, home, hscore, season, gid) %>% mutate(ha = 'away')
dh = d %>% select(date, home, hscore, away, ascore, season, gid) %>% mutate(ha = 'home')
colnames(da) = c('date', 'team', 'score',  'opp', 'opp.score', 'season', 'gid', 'ha')
colnames(dh) = c('date', 'team', 'score',  'opp', 'opp.score', 'season', 'gid', 'ha')
dd = bind_rows(da, dh) %>% 
  arrange(date, gid)
head(dd)


```

## Distance matrix grid plot

```{r fig.width = 6, fig.height = 7}

dg = dm %>% 
  left_join(tms %>% select(team, div), 
            by = c('prev.loc' = 'team')) %>%
  left_join(tms %>% select(team, div), 
            by = c('loc' = 'team'), 
            suffix = c('.1', '.2')) %>%
  mutate(prev.loc = factor(prev.loc, levels = rev(teams.order)), 
              loc = factor(     loc, levels =     teams.order))
head(dg)


title = "Miles Between Each Pair of Teams" 
g = ggplot(dg, 
           aes(x    = loc, 
               y    = prev.loc, 
               fill = miles))+
  geom_tile(linewidth   = 0.4, 
            show.legend = F, 
            color = pubdarkgray) +
  facet_grid(div.1 ~ div.2, 
             drop   = T, 
             scales = 'free') +
  scale_fill_gradient(low      = pubbackgray,
                      high     = pubred,
                      na.value = 'white',
                      oob      = squish) +
  labs(title = title,
       x = 'Home Team', 
       y = 'Away Team')
  
g %>% 
  pub(type = 'grid', 
      base_size = 36/4) + 
  theme(axis.text.x.top = element_text(angle = 90, 
                                       vjust = .5, 
                                       hjust = 0), 
        panel.spacing   = unit(10 * 1/1440 * 20 * 9/36, "in"))


```

## Total distance traveled by team

```{r}
dd = dd %>% 
  group_by(team) %>%
  mutate(loc = ifelse(ha == 'home', 
                      team, 
                      opp), 
         prev.loc = lag(loc), 
         prev.loc = ifelse(is.na(prev.loc), 
                           team, 
                           prev.loc)) %>%
  left_join(dm, by = c('loc', 'prev.loc'))

dd %>% 
  filter(team == 'PHI')
```

```{r}
dg = dd %>%
  group_by(team) %>%
  summarise(miles = sum(miles)) %>%
  arrange(miles) %>%
  left_join(tms, by = 'team') %>%
  mutate(team = factor(team, levels = team))
head(dg)
```

```{r}
max.x =  max(dg$miles) + 50000

title = "Distance travel by each team" 

g = ggplot(dg, 
           aes(x = miles, 
               y = team))+
  geom_bar(stat  = 'identity', 
           aes(x = max.x), 
           color = NA, 
           fill  = publightgray, 
           width = 0.8)+ ## optional background bars. 
  geom_bar(stat  = 'identity', 
           fill  = pubred, 
           color = NA, 
           width = 0.8)+ 
  geom_text(aes(label =  comma(miles)), 
            hjust = -0.1)+ ## optional numbers with reasonable number of digits
  facet_wrap(~div, 
             scales = 'free', 
             ncol = 3)+
  labs(title = title,
       x = 'Miles Traveled', ## Optional. 
       y = NULL)

g %>% 
  pub(type = 'bar', 
      xlim = c(0, max.x))
```


## Travel distance and game outcomes

Now let's see if there is any relationship between points scored and miles. 

```{r}
title = "Points Scored vs Miles" 
g = ggplot(dd, 
           aes(x = miles, 
               y = score))+
  geom_jitter(height = 0, 
              width  = 50, 
              alpha  = 0.5)+
  geom_smooth(method = 'lm', 
              color  = pubred)+
  labs(title = title,
       x = 'Miles Traveled From Previous Game',
       y = 'Points Scored')

g %>% 
  pub(xlim = c(0, 3000), 
      ylim = c(70, 160))
```

This relationship looks slightly decreasing but isn't that strong. 

## Travel paths

```{r}
ll = read.csv('data/nba.venue.lat.long.csv')

ll = ll %>%
  select(team, lat, long) %>%
  filter(!team %in% c('SEA', 'LV'))
head(ll)

dd = dd %>% 
  left_join(ll, by = c('prev.loc' = 'team')) %>%
  left_join(ll, by = c('loc'      = 'team'), 
            suffix = c('.prev', ''))

title = "NBA Travel Routes"

g = ggplot(dd, 
           aes(x = long, 
               y = lat)) +
  borders('state', 
          colour = publightgray) +
  geom_segment(aes(xend = long.prev, 
                   yend = lat.prev), 
               alpha = 0.03)+
  geom_point(data  = ll, 
             color = pubred)+
  geom_text(data = ll, 
            aes(label = team), 
            hjust = rep(c(-0.1, 1.1), 15))+
  labs(title = title)

g %>% 
  pub(type = 'map')
```

## Travel paths weighted by trips

```{r fig.width = 7, fig.height = 5}
dg = dd %>%
  group_by(prev.loc, loc, 
           lat     , long, 
           lat.prev, long.prev) %>%
  summarise(Trips = n())

title = "NBA Travel Routes Weighted By Number of Trips"

g = ggplot(dg, 
           aes(x = long, 
               y = lat))+
  borders('state', 
          colour = publightgray)+
  geom_segment(aes(xend = long.prev, 
                   yend = lat.prev, 
                   size = Trips), 
               alpha = 0.3)+
  geom_point(data  = ll, 
             color = pubred)+
  geom_text(data   = ll, 
            aes(label = team), 
            hjust  = rep(c(-0.1, 1.1), 15))+
  labs(title = title, 
       size  = 'Trips')

  
g %>% 
  pub(type = 'map') + 
  scale_size(range = c(.5,3)) +
  theme(legend.title = element_text(size = 36/3, 
                                    vjust = 0.5,
                                    margin = margin(0, 0, 0, 
                                                    150/72/3, 
                                                    'in')))
```

## Travel matrix

Let's show the number of times that teams travel from one city to another during the season. 

```{r}
dr = dg %>%
  ungroup() %>%
  filter(prev.loc != loc) %>%
  complete(prev.loc, 
           loc, 
           fill = list(Trips = 0)) %>%
  left_join(tms, by = c('prev.loc' = 'team')) %>%
  left_join(tms, by = c('loc'      = 'team'), 
            suffix = c('.prev', '')) %>%
  mutate(loc      = factor(loc     , levels =     teams.order), 
         prev.loc = factor(prev.loc, levels = rev(teams.order)))
head(dr,2)
```

```{r fig.height = 7, fig.width = 6}
title = "Number of Trips Between Each Pair of Cities" 

g = ggplot(dr, 
           aes(x = loc, 
               y = prev.loc, 
               fill = Trips))+
  geom_tile(linewidth = 0.4, 
            show.legend = T, 
            color = pubdarkgray) + 
  facet_grid(div.prev ~ div, 
             scales = 'free')+ ## now faceting by division
  scale_fill_gradient(low    = pubbackgray,
                      high   = pubred,
                      na.value = 'white',
                      oob    = squish, 
                      breaks = c(0,25), 
                      guide  = guide_colorbar(frame.colour = pubdarkgray))+
  labs(title = title,
       x     = 'Destination', 
       y     = 'Origin', 
       fill  = 'Number of Trips')

g %>% 
  pub(type = 'grid', 
      base_size = 36/4) +
  theme(axis.text.x.top = element_text(angle = 90, 
                                       vjust = .5, 
                                       hjust = 0), 
        panel.spacing     = unit(1/72*10/4, "in"), 
        legend.key.width  = unit(1/72*36/4, "in"),
        legend.title = element_text(size   = 36/4, 
                                    vjust  = 0.5, 
                                    margin = margin(0,  1/72*20/4,
                                                    0, -1/72*30/4, 
                                                    'in')))


```

Often an east coast team that goes on a road trip to the west coast will play several games against west coast teams. Specifically, they will play LAC and LAL in consecutive games. The same is true when teams travel to Florida and play consecutive games at MIA and ORL. 

We emphasize that this is not the same as the schedule matrix, which shows games played between each pair of teams, regardless of the travel routes they took to get to the game. In the travel matrix above, when *any* of the 30 teams in the league travel between MIA and ORL, or ORL and MIA, it counts as a travel route between those two cities.

